Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2019-02-13 multi_body_interactions/papers/theory/reviews_NJP/referee_response_1.tex

  We thank the referee for their careful reading of our manuscript,
  and for pointing out that our discussion in the paragraph between
  Eqs.~(8) and (9) is poorly worded.  The paragraph in question
  appears on page 7, lines 20--47:

  \red{We are interested in the ground and low-lying orbital
    eigenstates of several interacting atoms on a single lattice site,
    as such states can be readily prepared and coherently addressed in
    current experiments with ultracold atoms [34].  Although these
    experiments operate at temperatures sufficiently low for
    negligible thermal occupation of excited motional states, simply
    neglecting excited single-particle motional states fails to
    reproduce the observed spectra of multiply-occupied lattice sites.
    The reason for this failure lies in the fact that the bare
    two-body Hamiltonian $H_{\t{int}}$ in (7) is not diagonal in the
    basis of single-particle motional states.  Even at zero
    temperature, therefore, the presence of excited motional states
    modifies many-body atomic energy spectra when the magnitude of
    off-diagonal elements of $H_{\t{int}}$ approach the scale of
    low-lying single-particle motional state energy gaps.  To address
    this problem, we develop a low-energy effective theory for
    ultracold two-level fermionic atoms on a lattice.  As our work is
    motivated by progress in experiments which initialize all atoms in
    identical (i.e.~ground) orbital and motional states, to simplify
    our theory we assume that any $N$ atoms on a single lattice site
    occupy $N$ nuclear spin states.  Multiple occupation of a single
    nuclear spin state on the same lattice site is initially forbidden
    by fermionic statistics, and subsequently cannot be violated in
    the absence of tunneling and hyperfine coupling.}

  We have re-written this paragraph in its entirety to walk through
  the arguments and explanations therein more carefully, replacing it
  with the following text:

  \green{Current experiments with ${}^{87}$Sr can prepare up to five
    atoms in the same (ground) orbital state on a single lattice site,
    and coherently address states with a single orbital excitation per
    lattice site [34].  At ultracold temperatures well below the
    non-interacting motional excitation energies
    $\Delta_n\equiv E_n-E_0$ for $n>0$, atoms only occupy their
    motional ground state in the lattice.  For this reason, it is
    common to map the description of these atoms onto a single-band
    Hubbard model that captures all dynamics within the subspace of
    motional ground states of non-interacting atoms, i.e.~with
    wavefunctions $\phi_{i,0}$.  Interactions, however, modify atoms'
    motional ground-state wavefuctions.  The true motional ground
    state of a collection of interacting atoms is then an admixture of
    the non-interacting motional eigenstates, and a naive Hubbard
    model that assumes atomic wavefunctions $\phi_{i,0}$ will fail to
    reproduce the interacting atoms' orbital excitation spectrum.
    Formally, corrections to the spectrum of interacting atoms can be
    accounted for by a perturbative treatment of far-off-resonant
    terms in the interaction Hamiltonian of (7) that create atoms in
    excited motional states,
    e.g.~$\sim\c_{n\mu s}^\dag \c_{0,\nu t}^\dag \c_{0,\nu r}
    \c_{0,\mu q}$ with $n>0$.  These corrections can be understood
    through interaction-induced {\it virtual} occupation of higher
    bands (i.e.~excited motional states), which becomes relevant as
    more atoms occupy the same lattice site, such that their
    interaction energy becomes non-negligible compared to the motional
    excitation energies $\Delta_n$.}

### 2019-02-13 multi_body_interactions/papers/theory/paper.tex
changed instance of 'single-atom' to 'non-interacting', and added one revision

Current experiments with ${}^{87}$Sr can prepare up to five atoms in
the same (ground) orbital state on a single lattice site, and
coherently address states with a single orbital excitation per lattice
site \cite{goban2018emergence}.  At ultracold temperatures well below
the single-atom motional excitation energies $\Delta_n\equiv E_n-E_0$
for $n>0$, atoms only occupy their motional ground state in the
lattice.  For this reason, it is common to map the description of
these atoms onto a single-band Hubbard model that captures all
dynamics within the subspace of motional ground states of
non-interacting atoms, i.e.~with single-atom wavefunctions
$\phi_{i,0}$.  Interactions, however, modify atoms' motional
ground-state wavefuctions.  The true motional ground state of a
collection of interacting atoms is then an admixture of the
non-interacting motional eigenstates, and a naive Hubbard model that
assumes atomic wavefunctions $\phi_{i,0}$ will fail to reproduce the
interacting atoms' orbital excitation spectrum.  Formally, corrections
to the spectrum of interacting atoms can be accounted for by a
perturbative treatment of far-off-resonant terms in the interaction
Hamiltonian of \eqref{eq:H_int} that create atoms in excited motional
states,
e.g.~$\sim\c_{n\mu s}^\dag \c_{0,\nu t}^\dag \c_{0,\nu r} \c_{0,\mu
  q}$ with $n>0$.  These corrections can be understood through
interaction-induced {\it virtual} occupation of higher bands
(i.e.~excited motional states), which becomes relevant as more atoms
occupy the same lattice site, such that their interaction energy
becomes non-negligible compared to the motional excitation energies
$\Delta_n$.

@@

\begin{figure}
  \centering
  \subfloat{\includegraphics{sweep_V54_O50.pdf}}
  \subfloat{\includegraphics{sweep_expt_V54.pdf}}
  \caption{\footnotesize Population (in arbitrary units) of the
    excited ${}^3P_0$ orbital state of ${}^{87}$Sr atoms in a uniform
    mixture of nuclear spins.  Atoms are prepared in the ground state
    of a lattice with depth $\U=54 E_{\t{R}}$, where
    $E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon
    recoil energy of the atoms, and interrogated by a laser with Rabi
    frequency $\Omega_I=50\times2\pi~\t{Hz}$ for a time
    $t=\pi/\Omega_I$.  ({\bf Left}) Populations predicted by the
    low-energy effective theory (with $s$-wave scattering parameters
    retrieved from ref.~\cite{zhang2014spectroscopic}), averaged over
    all nuclear spin combinations of $N\in\set{1,\cdots,5}$ atoms per
    lattice site for a fixed total atom number.  ({\bf Right})
    Experimental measurements of ${}^3P_0$ populations retrieved from
    ref.~\cite{goban2018emergence}, with Lorentzian fits to each peak
    as a visual guide.  Resonance peaks are identified by the
    many-body orbital states which are excited at the peak.
    \label{fig:sweep}}
\end{figure}

We now consider samples of ${}^{87}$Sr atoms in a uniform mixture of
nuclear spins $\mu\in\set{-I,\cdots,I}$ prepared in motional ground
states of a rectangular lattice with depths
$\U=\p{\U_\x,\U_\y,\U_\z}=\p{41,55,69} E_{\t{R}}$, where
$E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon recoil
energy of the atoms.  Such samples can be prepared in experiments
which can vary the single-site occupation number $N$, and which can
control for the total number of atoms that are addressed by an
external interrogation laser.  Figure \ref{fig:sweep} shows the
population of the excited ${}^3P_0$ orbital state when these atoms are
interrogated for a time $t=\pi/\Omega_I$ by a laser with Rabi
frequency $\Omega_I=50\times2\pi~\t{Hz}$ (i.e.~for individual atoms
with nuclear spin $\mu=I$) and detuning $\Delta$ from the single-atom
${}^1S_0\to{}^3P_0$ orbital excitation energy.  The ${}^3P_0$
population peaks when the laser detuning $\Delta$ is equal to the
many-body excitation energy $\Delta_{NX}\equiv E_{NX}-E_{N,0}$ for
$X\in\set{+,-}$, as this is precisely when the on-resonance condition
$\delta=0$ is satisfied in the many-body Rabi Hamiltonian $H_{\N X}$
in \eqref{eq:H_NX}.  Due to experimental uncertainties which vary with
single-site occupation number $N$, the heights of experimental peaks
in figure \ref{fig:sweep} are not well-calibrated between different
values of $N$.  Nonetheless, figure \ref{fig:sweep} exhibits
signatures of larger asymmetric-state Rabi frequencies than
symmetric-state ones in the form of higher asymmetric-state peaks for
fixed $N$.

>>>>

Current experiments with ${}^{87}$Sr can prepare up to five atoms in
the same (ground) orbital state on a single lattice site, and
coherently address states with a single orbital excitation per lattice
site \cite{goban2018emergence}.  At ultracold temperatures well below
the non-interacting motional excitation energies
$\Delta_n\equiv E_n-E_0$ for $n>0$, atoms only occupy their motional
ground state in the lattice.  For this reason, it is common to map the
description of these atoms onto a single-band Hubbard model that
captures all dynamics within the subspace of motional ground states of
non-interacting atoms, i.e.~with wavefunctions $\phi_{i,0}$.
Interactions, however, modify atoms' motional ground-state
wavefuctions.  The true motional ground state of a collection of
interacting atoms is then an admixture of the non-interacting motional
eigenstates, and a naive Hubbard model that assumes atomic
wavefunctions $\phi_{i,0}$ will fail to reproduce the interacting
atoms' orbital excitation spectrum.  Formally, corrections to the
spectrum of interacting atoms can be accounted for by a perturbative
treatment of far-off-resonant terms in the interaction Hamiltonian of
\eqref{eq:H_int} that create atoms in excited motional states,
e.g.~$\sim\c_{n\mu s}^\dag \c_{0,\nu t}^\dag \c_{0,\nu r} \c_{0,\mu
  q}$ with $n>0$.  These corrections can be understood through
interaction-induced {\it virtual} occupation of higher bands
(i.e.~excited motional states), which becomes relevant as more atoms
occupy the same lattice site, such that their interaction energy
becomes non-negligible compared to the motional excitation energies
$\Delta_n$.

@@

\begin{figure}
  \centering
  \subfloat{\includegraphics{sweep_V54_O50.pdf}}
  \subfloat{\includegraphics{sweep_expt_V54.pdf}}
  \caption{\footnotesize Population (in arbitrary units) of the
    excited ${}^3P_0$ orbital state of ${}^{87}$Sr atoms in a uniform
    mixture of nuclear spins.  Atoms are prepared in the ground state
    of a lattice with depth $\U=54 E_{\t{R}}$, where
    $E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon
    recoil energy of the atoms, and interrogated by a laser with Rabi
    frequency $\Omega_I=50\times2\pi~\t{Hz}$ for a time
    $t=\pi/\Omega_I$.  ({\bf Left}) Populations predicted by the
    low-energy effective theory (with $s$-wave scattering parameters
    retrieved from ref.~\cite{zhang2014spectroscopic}), averaged over
    all nuclear spin combinations of $N\in\set{1,\cdots,5}$ atoms per
    lattice site for a fixed total atom number.  ({\bf Right})
    Experimental measurements of ${}^3P_0$ populations retrieved from
    ref.~\cite{goban2018emergence}, with Lorentzian fits to each peak
    as a visual guide.  Resonance peaks are identified by the
    many-body orbital states which are excited at the
    peak.}
  \label{fig:sweep}
\end{figure}

We now consider samples of ${}^{87}$Sr atoms in a uniform mixture of
nuclear spins $\mu\in\set{-I,-I+1,\cdots,I}$ prepared in motional
ground states of a rectangular lattice with depths
$\U=\p{\U_\x,\U_\y,\U_\z}=\p{41,55,69} E_{\t{R}}$, where
$E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon recoil
energy of the atoms.  Such samples can be prepared in experiments
which can vary the single-site occupation number $N$, and which can
control for the total number of atoms that are addressed by an
external interrogation laser.  Figure \ref{fig:sweep} shows the
population of the excited ${}^3P_0$ orbital state when these atoms are
interrogated for a time $t=\pi/\Omega_I$ by a laser with Rabi
frequency $\Omega_I=50\times2\pi~\t{Hz}$ (i.e.~for individual atoms
with nuclear spin $\mu=I$) and detuning $\Delta$ from the single-atom
${}^1S_0\to{}^3P_0$ orbital excitation energy.  The ${}^3P_0$
population peaks when the laser detuning $\Delta$ is equal to the
many-body excitation energy $\Delta_{NX}\equiv E_{NX}-E_{N,0}$ for
$X\in\set{+,-}$, as this is precisely when the on-resonance condition
$\delta=0$ is satisfied in the many-body Rabi Hamiltonian $H_{\N X}$
in \eqref{eq:H_NX}.  Due to experimental uncertainties which vary with
single-site occupation number $N$, the heights of experimental peaks
in figure \ref{fig:sweep} are not well-calibrated between different
values of $N$.  Nonetheless, figure \ref{fig:sweep} exhibits
signatures of larger asymmetric-state Rabi frequencies than
symmetric-state ones in the form of higher asymmetric-state peaks for
fixed $N$.

### 2019-02-13 multi_body_interactions/papers/theory/reviews_NJP/revision_summary.tex
changed instance of 'single-atom' to 'non-interacting', and added one revision

  \green{Current experiments with ${}^{87}$Sr can prepare up to five
    atoms in the same (ground) orbital state on a single lattice site,
    and coherently address states with a single orbital excitation per
    lattice site [34].  At ultracold temperatures well below the
    single-atom motional excitation energies $\Delta_n\equiv E_n-E_0$
    for $n>0$, atoms only occupy their motional ground state in the
    lattice.  For this reason, it is common to map the description of
    these atoms onto a single-band Hubbard model that captures all
    dynamics within the subspace of motional ground states of
    non-interacting atoms, i.e.~with single-atom wavefunctions
    $\phi_{i,0}$.  Interactions, however, modify atoms' motional
    ground-state wavefuctions.  The true motional ground state of a
    collection of interacting atoms is then an admixture of the
    non-interacting motional eigenstates, and a naive Hubbard model
    that assumes atomic wavefunctions $\phi_{i,0}$ will fail to
    reproduce the interacting atoms' orbital excitation spectrum.
    Formally, corrections to the spectrum of interacting atoms can be
    accounted for by a perturbative treatment of far-off-resonant
    terms in the interaction Hamiltonian of (7) that create atoms in
    excited motional states,
    e.g.~$\sim\c_{n\mu s}^\dag \c_{0,\nu t}^\dag \c_{0,\nu r}
    \c_{0,\mu q}$ with $n>0$.  These corrections can be understood
    through interaction-induced {\it virtual} occupation of higher
    bands (i.e.~excited motional states), which becomes relevant as
    more atoms occupy the same lattice site, such that their
    interaction energy becomes non-negligible compared to the motional
    excitation energies $\Delta_n$.}

@@

@@

\item For clarity, we have replaced a few instances of the text
  ``\red{single-particle}'' by ``\green{non-interacting}'' when
  referring to motional ground states of non-interacting particles;
  specifically, we have done so on page 9, line 50; on page 13, line
  5; and on page 35, line 26.

>>>>

  \green{Current experiments with ${}^{87}$Sr can prepare up to five
    atoms in the same (ground) orbital state on a single lattice site,
    and coherently address states with a single orbital excitation per
    lattice site [34].  At ultracold temperatures well below the
    non-interacting motional excitation energies
    $\Delta_n\equiv E_n-E_0$ for $n>0$, atoms only occupy their
    motional ground state in the lattice.  For this reason, it is
    common to map the description of these atoms onto a single-band
    Hubbard model that captures all dynamics within the subspace of
    motional ground states of non-interacting atoms, i.e.~with
    wavefunctions $\phi_{i,0}$.  Interactions, however, modify atoms'
    motional ground-state wavefuctions.  The true motional ground
    state of a collection of interacting atoms is then an admixture of
    the non-interacting motional eigenstates, and a naive Hubbard
    model that assumes atomic wavefunctions $\phi_{i,0}$ will fail to
    reproduce the interacting atoms' orbital excitation spectrum.
    Formally, corrections to the spectrum of interacting atoms can be
    accounted for by a perturbative treatment of far-off-resonant
    terms in the interaction Hamiltonian of (7) that create atoms in
    excited motional states,
    e.g.~$\sim\c_{n\mu s}^\dag \c_{0,\nu t}^\dag \c_{0,\nu r}
    \c_{0,\mu q}$ with $n>0$.  These corrections can be understood
    through interaction-induced {\it virtual} occupation of higher
    bands (i.e.~excited motional states), which becomes relevant as
    more atoms occupy the same lattice site, such that their
    interaction energy becomes non-negligible compared to the motional
    excitation energies $\Delta_n$.}

@@

\item In figure 2 (page 22), all multiplicities (i.e.~values on the
  vertical axes) are integer-valued.  We have therefore adjusted the
  vertical axis on the top left panel of this figure, such that now
  only integer values are marked on this axis.

\item For slight clarification, we have changed the text (page 22,
  line 51):

  \red{... nuclear spins $\mu\in\set{-I,\cdots,I}$ ...}

  \green{... nuclear spins $\mu\in\set{-I,-I+1,\cdots,I}$ ...}

@@

\item For clarity, we have replaced a few instances of the text
  ``\red{single-particle}'' or ``\red{single-atom}'' by
  ``\green{non-interacting}'' when referring to motional ground states
  of non-interacting particles; specifically, we have done so on page
  9, line 50; on page 11, line 22; on page 13, line 5; and on page 35,
  line 26.
